### 2017-11-07 submission/manuscript.Rmd
method edits

***Obtaining Data Sets:***
Studies used for this meta-analysis were identified through the review articles written by Keku, et al. and Vogtmann, et al. [@keku_gastrointestinal_2015; @vogtmann_epidemiologic_2016]. All studies were included that used tissue or feces as their sample source for 16S rRNA gene sequencing analysis. Studies using either 454 or Illumina sequencing technology were included. Only data sets that had the raw sequences available for analysis were included. Some studies did not have publically available raw sequences or did not have meta data in which the authors were able to share. After this filtering step the following studies remained: Ahn [@ahn_human_2013], Baxter [@baxter_microbiota-based_2016], Brim [@brim_microbiome_2013], Burns [@burns_virulence_2015], Chen [@chen_human_2012], Dejea [@dejea_microbiota_2014], Flemer [@flemer_tumour-associated_2017], Geng [@geng_diversified_2013], Hale [@hale_shifts_2017], Kostic [@kostic_genomic_2012], Lu [@lu_mucosal_2016], Sanapareddy [@sanapareddy_increased_2012], Wang [@wang_structural_2012], Weir [@weir_stool_2013], and Zeller [@zeller_potential_2014]. The Zackular [@zackular_human_2014] study was not included becasue the 90 individuals analyzed within the study are contained within the larger Baxter study. The Kostic study was not used since after sequence processing all the case samples did not have more than 100 sequences remaining. This left a total of 13 studies in which complete analysis could be completed.

@@

***Sequence Processing:***
For the majority of studies raw sequences were downloaded from the SRA (ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/) and metadata was obtained from the following website: http://www.ncbi.nlm.nih.gov/Traces/study/ by searching the respective accession number of the study. Of the studies that did not have sequences and meta data on the SRA one study had the data stored on DBGap [@ahn_human_2013] and four studies the data was obtained directly from the authors [@flemer_tumour-associated_2017; @hale_shifts_2017; @sanapareddy_increased_2012; @weir_stool_2013]. Each study was processed using the mothur (v1.39.3) software program [@schloss_introducing_2009]. Where possible quality filtering utilized the default methods used in mothur for either 454 or Illumina based sequencing. If it was not possible to use these defaults the author stated quality cut-offs were used instead. Chimeras were identifed and removed using the VSEARCH [@rognes_vsearch_2016] program and *de novo* OTU clustering at 97% similarity using the OptiClust algorithm [@westcott_opticlust_2017] was utilized.

***Statistical Analysis:***
All statistical analysis after sequence processing utilized the R software package (v`r paste(version$major, ".", version$minor, sep = "")`). For the alpha diversity analysis values were power transformed using the rcompanion (v`r paste(packageVersion("rcompanion")[1])`) package and then Z-score normalized using the car (v`r paste(packageVersion("car")[1])`) package. Testing for alpha diversity differences utilized linear mixed-effect models created using the lme4 (v`r paste(packageVersion("lme4"))[1]`) package to correct for both study and variable region effect in the diversity measures when analyzing colorectal cancer groups. Relative Risk was analyzed using both the epiR (v`r paste(packageVersion("epiR"))[1]`) and metafor (v`r paste(packageVersion("metafor"))[1]`) packages. Relative risk significance testing utilized the chi-squred test. Beta-diversity differences utilized a Bray-Curtis distance matrix and PERMANOVA executed with the vegan (v`r paste(packageVersion("vegan"))[1]`) package. Random Forest models were built using both the caret (v`r paste(packageVersion("caret"))[1]`) and randomForest (v`r paste(packageVersion("randomForest"))[1]`) packages. Random Forest testing of the obtained AUC versus a random model AUC utilized T-tests. Power analysis and estimations were made using the pwr (v`r paste(packageVersion("pwr"))[1]`) and statmod (v`r paste(packageVersion("statmod"))[1]`) packages. All figures were created using both ggplot2 (v`r paste(packageVersion("ggplot2"))[1]`) and gridExtra (v`r paste(packageVersion("gridExtra"))[1]`) packages.

***Study Analysis Overview:***
Alpha diversity was first assessed for differences between controls and adenoma versus cancer and controls versus adenoma. We analyzed the data using linear mixed-effect models, and relative risk. Beta-diversity was then assessed for each inidividual study. Next, four specific CRC-associated genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, and *Porphyromonas*) were assessed for differences in relative risk. We then built Random Forest models based on all genera or the select CRC-associated genera. The models were trained on one study then tested on the remaining studies for every study. The data was split between feces and tissue samples. Within the tissue groups the data was further divided between matched and unmatched tissue samples. Both prediction for adenoma and carcinoma were tested. This same approach was then applied at the OTU level with the exception that instead of testing on the other studies a 10-fold cross validation was utilized and 100 different models were created based on random 80/20 splitting of the data to generate a range of expected AUCs. For OTU based models the CRC Associated Genera included all OTUs that had a taxonomic classification to *Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*. The power of each study was assessed for and effect size ranging from 1% to 30%. An estimated sample n for these effect sizes was also generated based on 80% power.

>>>>

***Obtaining Data Sets:***
Studies used for this meta-analysis were identified through the review articles written by Keku, et al. and Vogtmann, et al. [@keku_gastrointestinal_2015; @vogtmann_epidemiologic_2016]. Additional studies not mentioned in the reviews were obtained from the authors knowledge of the literature. All studies were included that used tissue or feces as their sample source for 16S rRNA gene sequencing analysis. Studies using either 454 or Illumina sequencing technology were included. Only data sets that had sequences available for analysis were included. Some studies did not have publically available sequences or did not have meta data in which the authors were able to share. After these filtering steps the following studies remained: Ahn [@ahn_human_2013], Baxter [@baxter_microbiota-based_2016], Brim [@brim_microbiome_2013], Burns [@burns_virulence_2015], Chen [@chen_human_2012], Dejea [@dejea_microbiota_2014], Flemer [@flemer_tumour-associated_2017], Geng [@geng_diversified_2013], Hale [@hale_shifts_2017], Kostic [@kostic_genomic_2012], Lu [@lu_mucosal_2016], Sanapareddy [@sanapareddy_increased_2012], Wang [@wang_structural_2012], Weir [@weir_stool_2013], and Zeller [@zeller_potential_2014]. The Zackular [@zackular_human_2014] study was not included becasue the 90 individuals analyzed within the study are contained within the larger Baxter study. The Kostic study was not used since after sequence processing all the case samples did not have more than 100 sequences remaining. This left a total of 14 studies in which complete analysis could be completed.

@@

***Sequence Processing:***
For the majority of studies raw sequences were downloaded from the SRA (ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/) and metadata was obtained from the following website: http://www.ncbi.nlm.nih.gov/Traces/study/ by searching the respective accession number of the study. Of the studies that did not have sequences and meta data on the SRA one study had the data stored on DBGap [@ahn_human_2013] and for four studies the data was obtained directly from the authors [@flemer_tumour-associated_2017; @hale_shifts_2017; @sanapareddy_increased_2012; @weir_stool_2013]. Each study was processed using the mothur (v1.39.3) software program [@schloss_introducing_2009]. Where possible quality filtering utilized the default methods used in mothur for either 454 or Illumina based sequencing. If it was not possible to use these defaults the author stated quality cut-offs were used instead. Chimeras were identifed and removed using the VSEARCH [@rognes_vsearch_2016] program and *de novo* OTU clustering at 97% similarity using the OptiClust algorithm [@westcott_opticlust_2017] was utilized.

***Statistical Analysis:***
All statistical analysis after sequence processing utilized the R software package (v`r paste(version$major, ".", version$minor, sep = "")`). For the alpha diversity analysis values were power transformed using the rcompanion (v`r paste(packageVersion("rcompanion")[1])`) package and then Z-score normalized using the car (v`r paste(packageVersion("car")[1])`) package. Testing for alpha diversity differences utilized linear mixed-effect models created using the lme4 (v`r paste(packageVersion("lme4"))[1]`) package to correct for study and variable region effects in feces and study, variable region, and individual effects in tissue. Relative Risk was analyzed using both the epiR (v`r paste(packageVersion("epiR"))[1]`) and metafor (v`r paste(packageVersion("metafor"))[1]`) packages. Relative risk significance testing utilized the chi-squred test. Beta-diversity differences utilized a Bray-Curtis distance matrix and PERMANOVA executed with the vegan (v`r paste(packageVersion("vegan"))[1]`) package. Random Forest models were built using both the caret (v`r paste(packageVersion("caret"))[1]`) and randomForest (v`r paste(packageVersion("randomForest"))[1]`) packages. Random Forest testing of the obtained AUC versus a random model AUC utilized T-tests. Power analysis and estimations were made using the pwr (v`r paste(packageVersion("pwr"))[1]`) and statmod (v`r paste(packageVersion("statmod"))[1]`) packages. All figures were created using both ggplot2 (v`r paste(packageVersion("ggplot2"))[1]`) and gridExtra (v`r paste(packageVersion("gridExtra"))[1]`) packages.

***Study Analysis Overview:***
Alpha diversity was first assessed for differences between controls, adenoma, and carcinoma. We analyzed the data using linear mixed-effect models and relative risk. Beta-diversity was then assessed for each inidividual study. Next, four specific CRC-associated genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, and *Porphyromonas*) were assessed for differences in relative risk. We then built Random Forest models based on all genera or the select CRC-associated genera. The models were trained on one study then tested on the remaining studies for every study. The data was split between feces and tissue samples. Within the tissue groups the data was further divided between matched and unmatched tissue samples. Where applicable for each study predictions for adenoma and carcinoma were tested. This same approach was then applied at the OTU level with the exception that instead of testing on the other studies a 10-fold cross validation was utilized and 100 different models were created based on random 80/20 splitting of the data to generate a range of expected AUCs. For OTU based models the CRC associated genera included all OTUs that had a taxonomic classification to *Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*. The power of each study was assessed for an effect size ranging from 1% to 30%. An estimated sample n for these effect sizes was also generated based on 80% power.

### 2017-11-07 submission/manuscript.tex
method edits

\textbf{\emph{Obtaining Data Sets:}} Studies used for this meta-analysis
were identified through the review articles written by Keku, et al. and
Vogtmann, et al. {[}26,27{]}. All studies were included that used tissue
or feces as their sample source for 16S rRNA gene sequencing analysis.
Studies using either 454 or Illumina sequencing technology were
included. Only data sets that had the raw sequences available for
analysis were included. Some studies did not have publically available
raw sequences or did not have meta data in which the authors were able
to share. After this filtering step the following studies remained: Ahn
{[}21{]}, Baxter {[}24{]}, Brim {[}17{]}, Burns {[}22{]}, Chen {[}14{]},
Dejea {[}19{]}, Flemer {[}13{]}, Geng {[}25{]}, Hale {[}12{]}, Kostic
{[}28{]}, Lu {[}16{]}, Sanapareddy {[}20{]}, Wang {[}15{]}, Weir
{[}18{]}, and Zeller {[}23{]}. The Zackular {[}29{]} study was not
included becasue the 90 individuals analyzed within the study are
contained within the larger Baxter study. The Kostic study was not used
since after sequence processing all the case samples did not have more
than 100 sequences remaining. This left a total of 13 studies in which
complete analysis could be completed.

@@

\textbf{\emph{Sequence Processing:}} For the majority of studies raw
sequences were downloaded from the SRA
(\url{ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/})
and metadata was obtained from the following website:
\url{http://www.ncbi.nlm.nih.gov/Traces/study/} by searching the
respective accession number of the study. Of the studies that did not
have sequences and meta data on the SRA one study had the data stored on
DBGap {[}21{]} and four studies the data was obtained directly from the
authors {[}12,13,18,20{]}. Each study was processed using the mothur
(v1.39.3) software program {[}30{]}. Where possible quality filtering
utilized the default methods used in mothur for either 454 or Illumina
based sequencing. If it was not possible to use these defaults the
author stated quality cut-offs were used instead. Chimeras were
identifed and removed using the VSEARCH {[}31{]} program and \emph{de
novo} OTU clustering at 97\% similarity using the OptiClust algorithm
{[}32{]} was utilized.

\textbf{\emph{Statistical Analysis:}} All statistical analysis after
sequence processing utilized the R software package (v3.4.2). For the
alpha diversity analysis values were power transformed using the
rcompanion (v1.10.1) package and then Z-score normalized using the car
(v2.1.5) package. Testing for alpha diversity differences utilized
linear mixed-effect models created using the lme4 (v1.1.14) package to
correct for both study and variable region effect in the diversity
measures when analyzing colorectal cancer groups. Relative Risk was
analyzed using both the epiR (v0.9.87) and metafor (v2.0.0) packages.
Relative risk significance testing utilized the chi-squred test.
Beta-diversity differences utilized a Bray-Curtis distance matrix and
PERMANOVA executed with the vegan (v2.4.4) package. Random Forest models
were built using both the caret (v6.0.77) and randomForest (v4.6.12)
packages. Random Forest testing of the obtained AUC versus a random
model AUC utilized T-tests. Power analysis and estimations were made
using the pwr (v1.2.1) and statmod (v1.4.30) packages. All figures were
created using both ggplot2 (v2.2.1) and gridExtra (v2.3) packages.

\textbf{\emph{Study Analysis Overview:}} Alpha diversity was first
assessed for differences between controls and adenoma versus cancer and
controls versus adenoma. We analyzed the data using linear mixed-effect
models, and relative risk. Beta-diversity was then assessed for each
inidividual study. Next, four specific CRC-associated genera
(\emph{Fusobacterium}, \emph{Parvimonas}, \emph{Peptostreptococcus}, and
\emph{Porphyromonas}) were assessed for differences in relative risk. We
then built Random Forest models based on all genera or the select
CRC-associated genera. The models were trained on one study then tested
on the remaining studies for every study. The data was split between
feces and tissue samples. Within the tissue groups the data was further
divided between matched and unmatched tissue samples. Both prediction
for adenoma and carcinoma were tested. This same approach was then
applied at the OTU level with the exception that instead of testing on
the other studies a 10-fold cross validation was utilized and 100
different models were created based on random 80/20 splitting of the
data to generate a range of expected AUCs. For OTU based models the CRC
Associated Genera included all OTUs that had a taxonomic classification
to \emph{Fusobacterium}, \emph{Parvimonas}, \emph{Peptostreptococcus},
or \emph{Porphyromonas}. The power of each study was assessed for and
effect size ranging from 1\% to 30\%. An estimated sample n for these
effect sizes was also generated based on 80\% power.

>>>>

\textbf{\emph{Obtaining Data Sets:}} Studies used for this meta-analysis
were identified through the review articles written by Keku, et al. and
Vogtmann, et al. {[}26,27{]}. Additional studies not mentioned in the
reviews were obtained from the authors knowledge of the literature. All
studies were included that used tissue or feces as their sample source
for 16S rRNA gene sequencing analysis. Studies using either 454 or
Illumina sequencing technology were included. Only data sets that had
sequences available for analysis were included. Some studies did not
have publically available sequences or did not have meta data in which
the authors were able to share. After these filtering steps the
following studies remained: Ahn {[}21{]}, Baxter {[}24{]}, Brim
{[}17{]}, Burns {[}22{]}, Chen {[}14{]}, Dejea {[}19{]}, Flemer
{[}13{]}, Geng {[}25{]}, Hale {[}12{]}, Kostic {[}28{]}, Lu {[}16{]},
Sanapareddy {[}20{]}, Wang {[}15{]}, Weir {[}18{]}, and Zeller {[}23{]}.
The Zackular {[}29{]} study was not included becasue the 90 individuals
analyzed within the study are contained within the larger Baxter study.
The Kostic study was not used since after sequence processing all the
case samples did not have more than 100 sequences remaining. This left a
total of 14 studies in which complete analysis could be completed.

@@

\textbf{\emph{Sequence Processing:}} For the majority of studies raw
sequences were downloaded from the SRA
(\url{ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/})
and metadata was obtained from the following website:
\url{http://www.ncbi.nlm.nih.gov/Traces/study/} by searching the
respective accession number of the study. Of the studies that did not
have sequences and meta data on the SRA one study had the data stored on
DBGap {[}21{]} and for four studies the data was obtained directly from
the authors {[}12,13,18,20{]}. Each study was processed using the mothur
(v1.39.3) software program {[}30{]}. Where possible quality filtering
utilized the default methods used in mothur for either 454 or Illumina
based sequencing. If it was not possible to use these defaults the
author stated quality cut-offs were used instead. Chimeras were
identifed and removed using the VSEARCH {[}31{]} program and \emph{de
novo} OTU clustering at 97\% similarity using the OptiClust algorithm
{[}32{]} was utilized.

\textbf{\emph{Statistical Analysis:}} All statistical analysis after
sequence processing utilized the R software package (v3.4.2). For the
alpha diversity analysis values were power transformed using the
rcompanion (v1.10.1) package and then Z-score normalized using the car
(v2.1.5) package. Testing for alpha diversity differences utilized
linear mixed-effect models created using the lme4 (v1.1.14) package to
correct for study and variable region effects in feces and study,
variable region, and individual effects in tissue. Relative Risk was
analyzed using both the epiR (v0.9.87) and metafor (v2.0.0) packages.
Relative risk significance testing utilized the chi-squred test.
Beta-diversity differences utilized a Bray-Curtis distance matrix and
PERMANOVA executed with the vegan (v2.4.4) package. Random Forest models
were built using both the caret (v6.0.77) and randomForest (v4.6.12)
packages. Random Forest testing of the obtained AUC versus a random
model AUC utilized T-tests. Power analysis and estimations were made
using the pwr (v1.2.1) and statmod (v1.4.30) packages. All figures were
created using both ggplot2 (v2.2.1) and gridExtra (v2.3) packages.

\textbf{\emph{Study Analysis Overview:}} Alpha diversity was first
assessed for differences between controls, adenoma, and carcinoma. We
analyzed the data using linear mixed-effect models and relative risk.
Beta-diversity was then assessed for each inidividual study. Next, four
specific CRC-associated genera (\emph{Fusobacterium}, \emph{Parvimonas},
\emph{Peptostreptococcus}, and \emph{Porphyromonas}) were assessed for
differences in relative risk. We then built Random Forest models based
on all genera or the select CRC-associated genera. The models were
trained on one study then tested on the remaining studies for every
study. The data was split between feces and tissue samples. Within the
tissue groups the data was further divided between matched and unmatched
tissue samples. Where applicable for each study predictions for adenoma
and carcinoma were tested. This same approach was then applied at the
OTU level with the exception that instead of testing on the other
studies a 10-fold cross validation was utilized and 100 different models
were created based on random 80/20 splitting of the data to generate a
range of expected AUCs. For OTU based models the CRC associated genera
included all OTUs that had a taxonomic classification to
\emph{Fusobacterium}, \emph{Parvimonas}, \emph{Peptostreptococcus}, or
\emph{Porphyromonas}. The power of each study was assessed for an effect
size ranging from 1\% to 30\%. An estimated sample n for these effect
sizes was also generated based on 80\% power.
